### 2023-04-19 paper.Rmd
misc

Some simplifications in the computation of $\mathit{SSE_{kplusGeneralized}}$ can be arranged when ignoring the moment weights $\mathbf{w}$, e.g. by setting $w_t = 1$ ($\forall t = 1, \dots, T$). This simplification leads to the *unweighted sum approach* for k-plus anticlustering. In this case, the additional dimension in $\mathbf{\hat{X}}_{T \times N \times P}$, representing distribution moments can be dropped; instead, the tensor is flattened into a two dimensional matrix, where the variables representing the additional distribution moments are appended to the original data matrix $\mathbf{X}_{N \times P}$, resulting in the augmented matrix $\mathbf{\hat{X}'}_{N \times \mathit{TP}}$:

@@

Thus, when setting all the weights for all moments to 1, the k-plus criterion $\mathit{SSE_{kplusGeneralized}}$ again has the same form as the standard k-means criterion $\mathit{SSE}$, but it is computed on different data. Here, computing the k-plus criterion is done by first computing the additional $t$-moment variables, appending them as new columns to the original data matrix, and then computing $\mathit{SSE}$ on the augmented data. Using the unweighted sum approach, k-plus anticlustering therefore leaves the original k-means criterion unchanged.

Note that a problem may arise if the weights $\mathbf{w}$ are ignored when computing. Because the computation of the new $t$-moment variables involves squaring, their range of values is expected to be very different from the raw values $x_{ij}$ (e.g., see Figure 1). This imbalance can strongly and unpredictably influence the relative weight of $\mathit{SSE}$ and $\mathit{SSE_{Var}}$ when computing $\mathit{SSE_{kplus}}$. To ensure an equal weight of both criteria, all attributes---i.e., the original data as well as the new $t$-moment varibles---can be standardized before computing $\mathit{SSE_{kplusGeneralized}}$, e.g. via $z$-standardization.

@@

The simulation studies were implemented using the statistical programming language R [Version 4.2, @R-base]. The R package `anticlust` [Version 0.6.1, @R-anticlust; @papenberg2020] was used to optimize the anticlustering objectives. The R package `faux` [Version 1.1.0, @R-faux] was used to generate the multivariate normal data sets that were processed during the simulation. The R packages `dplyr` [Version 1.0.7, @R-dplyr], `ggplot2` [Version 3.3.5, @R-ggplot2], `papaja` [Version 0.1.1.9001, @R-papaja], `tidyr` [Version 1.1.3, @R-tidyr] and `DescTools` [Version 0.99.45, @R-desctools] were used for data analysis and result presentation. The simulation study is fully reproducible via code and data that has been made accessible in an accompanying OSF repository [@osf2023].

@@

In general, the weighting scheme for k-plus anticlustering can become highly complex as it might encompass an arbitrary number of distribution moments, and hence, an arbitrary number of weights. While an exhaustive evaluation of the influence of weights is beyond the scope of the current paper, a small scale simulation study was conducted to investigate the influence of weights for a selected number of moments. The target was the optimization of $\mathit{SSE_{kplusGeneralized}}$ with $T = 4$, i.e. minizimizing differences with regard to means, variances, skew and kurtosis. The two unweighted approaches were implemented with $\mathbf{w} = (1, 1, 1, 1)$. In one case, a standardization of all columns of $\mathbf{\hat{X}'}_{N \times \mathit{TP}}$ was implemented. That is, the original as well as the augmented variables were standardized before the criterion $\mathit{SSE_{kplusGeneralized}}$ was optimized. In the second case, the unstandardized matrix $\mathbf{\hat{X}'}_{N \times \mathit{TP}}$ was used. The four weighted approaches were implemented by giving a 10 times increased weight to one of the respective moments. That is, to compute $\mathit{SSE_{kplusGeneralized}}$ favoring similarity in means during the optimization process, I set $\mathbf{w} = (10, 1, 1, 1)$; the weights favoring similarity in variances, skewness and kurtosis, were defined as $\mathbf{w} = (1, 10, 1, 1)$, $\mathbf{w} = (1, 1, 10, 1)$, and $\mathbf{w} = (1, 1, 1, 10)$, repectively. The weighted approaches did not use standardization of the input data.

To compare the different weighting schemes, `r length(results)` data sets from were generated from a normal distribution ($M = 0$, $\mathit{SD} = 1$). The number of variables $P$ was determined randomly for each data set and varied between 1 and 5; the number of anticlusters $K$ was also determined randomly for each data set and varied between 2 and 4; the sample size $N$ randomly varied between 10 and 100 with the restriction that $N$ had to be divisible by $K$. For each data set, $\mathit{SSE_{kplusGeneralized}}$ was optimized using the local maximum algorithm, once for each of the six competing weighting schemes.

```{r}
final_tab_sim1 <- (Reduce("+", results) / length(results) * 100) |> round(2)
rownames(final_tab_sim1) <- c(
  "$\\mathbf{w} = (1, 1, 1, 1)$ / unstandardized",
  "$\\mathbf{w} = (1, 1, 1, 1)$ / standardized",
  "$\\mathbf{w} = (10, 1, 1, 1)$",
  "$\\mathbf{w} = (1, 10, 1, 1)$",
  "$\\mathbf{w} = (1, 1, 10, 1)$",
  "$\\mathbf{w} = (1, 1, 1, 10)$"
)
colnames(final_tab_sim1) <- c(
  "$\\mathit{SSE}$",
  "$\\mathit{SSE_{Var}}$", 
  "$\\mathit{SSE_{Skew}}$",
  "$\\mathit{SSE_{Kurtosis}}$",
  "Average"
)
note <- paste0("The results of Simulation 1 averaged across all ", length(results), " data sets. Cells indicate how well the maximization of the weighted $\\mathit{SSE_{kplusGeneralized}}$ fulfilled each of the four constituting criteria $\\mathit{SSE}$, $\\mathit{SSE_{Var}}$, $\\mathit{SSE_{Skew}}$, and $\\mathit{SSE_{Kurtosis}}$. Cell values represent relative performance, i.e. they were computed as the $\\mathit{SSE}$ that was attained by the given weighting scheme, divided by the best $\\mathit{SSE}$ that was obtained by any weighting scheme in a given simulation run.")

@@

Table 1 illustrates how well each of the six weighting schemes fulfilled each of the four constititung criteria, averaged across the `r length(results)` simulation runs. It is shown that for the weighted approaches, the criterion that received the highest weight was fulfilled to the highest degree, as should be expected. Interestingly, the unweighted approach including standardization had the best average performance across the four criteria. While the advantage over the other weighting schemes does not seem to be remarkable, this result held across an astonishing `r sum(unstd_best)` simulation runs (`r mean(unstd_best) * 100`%). Among the competing methods, the unweighted standardization approach therefore seems preferable. It is also a strong approach because it yielded the best results with regard to maximizing the standard $\mathit{SSE}$. Hence, the unweighted standardization approach implicitly gives high precedence to minimizing differences in means. This is a pleasant result because, as discussed earlier, for the higher order moment $\mathit{SSE}$s to be valid, it has to be ensured that the group means are close to the overall mean, which is best accomplished if the standard $\mathit{SSE}$ receives high priority.

@@

Figure 2 shows that with increasing $N$, k-means and the k-plus methods converged on the same level with regard to minimizing $\Delta \mathit{M}$. This is an important observation because it shows that k-plus anticlustering is capable of addressing multiple objectives at the same time---especially when $N$ increases. However, while k-plus anticlustering was capable of addressing multiple objectives simultaneously, increasing the number of optimization criteria made it more difficult to fulfill each single criterion: k-plus skew/kurtosis performed worse at minimizing $\Delta \mathit{M}$ and $\Delta \mathit{SD}$ than standard k-plus that only considered the mean and variances. However, as intended, k-plus skew/kurtosis was best at minimizing $\Delta \mathit{Skew}$ and $\Delta \mathit{Kurtosis}$ and at the same time, this k-plus objective maintained a comparably good level of addressing $\Delta \mathit{M}$ and $\Delta \mathit{SD}$, outperforming diversity anticlustering.

It is maybe surprising that k-plus skew/kurtosis does not seem to have performed much better than diversity anticlustering with regard to skewness and kurtosis, even though the method has been specifically tailored to these criteria. Figure 2 gives a more fine grained assessment of this observation: Surprisingly, for small $N$, k-plus skew/kurtosis was even worse than diversity anticlustering at minimizing $\Delta \mathit{Skew}$ and $\Delta \mathit{Kurtosis}$. However, with increasing $N$, it clearly outperformed diversity anticlustering. These results illustrate the cost of optimizing several criteria at the same time: For low $N$, k-plus skew/kurtosis was apparently preoccupied with fulfilling the objectives $\Delta \mathit{M}$ and $\Delta \mathit{SD}$. Increasing $N$ then facilitated to also address $\Delta \mathit{Skew}$ and $\Delta \mathit{Kurtosis}$.

>>>>

Some simplifications in the computation of $\mathit{SSE_{kplusGeneralized}}$ can be arranged when ignoring the moment weights $\mathbf{w}$, e.g. by setting $w_t = 1$ ($t = 1, \dots, T$). This simplification leads to the *unweighted sum approach* for k-plus anticlustering. In this case, the additional dimension in $\mathbf{\hat{X}}_{T \times N \times P}$, representing distribution moments can be dropped; instead, the tensor is flattened into a two dimensional matrix, where the variables representing the additional distribution moments are appended to the original data matrix $\mathbf{X}_{N \times P}$, resulting in the augmented matrix $\mathbf{\hat{X}'}_{N \times \mathit{TP}}$:

@@

Thus, when setting all the weights for all moments to 1, the k-plus criterion $\mathit{SSE_{kplusGeneralized}}$ again has the same form as the standard k-means criterion $\mathit{SSE}$, but it is computed on different data. Here, computing the k-plus criterion is done by first creating the additional $t$-moment variables, appending them as new columns to the original data matrix, and then computing $\mathit{SSE}$ on the augmented data. Using the unweighted sum approach, k-plus anticlustering therefore leaves the original k-means criterion unchanged.

Note that a problem may arise if the weights $\mathbf{w}$ are ignored when computing $\mathit{SSE_{kplusGeneralized}}$. Because the computation of the new $t$-moment variables involves exponentiation, their range of values is expected to be very different from the raw values $x_{ij}$ (e.g., see Figure 1). This imbalance can strongly and unpredictably influence the relative weight of the different moments when computing $\mathit{SSE_{kplusGeneralized}}$. To ensure an equal weight of all criteria, all attributes---i.e., the original data as well as the new $t$-moment variables---can be standardized before computing $\mathit{SSE_{kplusGeneralized}}$, e.g. via $z$-standardization. This approach will henceforth be referred to as the unweighted standardization approach for k-plus anticlustering.

@@

The simulation studies were implemented using the statistical programming language R [Version 4.2, @R-base]. The R package `anticlust` [Version 0.6.1, @R-anticlust; @papenberg2020] was used to optimize the anticlustering objectives. The R package `faux` [Version 1.1.0, @R-faux] was used to generate the multivariate normal data sets that were processed during Simulation 2. The R packages `dplyr` [Version 1.0.7, @R-dplyr], `ggplot2` [Version 3.3.5, @R-ggplot2], `papaja` [Version 0.1.1.9001, @R-papaja], `tidyr` [Version 1.1.3, @R-tidyr] and `DescTools` [Version 0.99.45, @R-desctools] were used for data analysis and result presentation. The simulation study is fully reproducible via code and data that has been made accessible in an accompanying OSF repository [@osf2023].

@@

In general, the weighting scheme for k-plus anticlustering can become highly complex as it might encompass an arbitrary number of distribution moments, and hence, an arbitrary number of weights. While an exhaustive evaluation of the influence of weights is beyond the scope of the current paper, a small scale simulation study was conducted to investigate the influence of weights for a selected number of moments. The target was the optimization of $\mathit{SSE_{kplusGeneralized}}$ with $T = 4$, i.e. minizimizing differences with regard to means, variances, skew and kurtosis. Two unweighted approaches (using standardization vs. not using standardization) were implemented with $\mathbf{w} = (1, 1, 1, 1)$. The unweighted standardization approach implemented a $z$-standardization of all columns---the original as well as the augmented variables---of $\mathbf{\hat{X}'}_{N \times \mathit{TP}}$ before the criterion $\mathit{SSE_{kplusGeneralized}}$ was optimized. In the other case, the unstandardized matrix $\mathbf{\hat{X}'}_{N \times \mathit{TP}}$ was used. The four weighted approaches were implemented by giving a tenfold increased weight to only one of the four moments. That is, to compute $\mathit{SSE_{kplusGeneralized}}$ favoring similarity in means during the optimization process, I set $\mathbf{w} = (10, 1, 1, 1)$; the weights favoring similarity in variances, skewness and kurtosis, were defined as $\mathbf{w} = (1, 10, 1, 1)$, $\mathbf{w} = (1, 1, 10, 1)$, and $\mathbf{w} = (1, 1, 1, 10)$, repectively. The weighted approaches did not use standardization of the input data.

To compare the different weighting schemes, `r length(results)` data sets from were generated from a normal distribution ($M = 0$, $\mathit{SD} = 1$). The number of variables $P$ was determined randomly for each data set and varied between 1 and 5; the number of anticlusters $K$ was also determined randomly for each data set and varied between 2 and 4; the sample size $N$ randomly varied between 20 and 100 with the restriction that $N$ had to be divisible by $K$. For each data set, $\mathit{SSE_{kplusGeneralized}}$ was optimized using the local maximum algorithm, once for each of the six competing weighting schemes.

```{r}
final_tab_sim1 <- round(Reduce("+", results) / length(results) * 100, 2)
rownames(final_tab_sim1) <- c(
  "$\\mathbf{w} = (1, 1, 1, 1)$ / unstandardized",
  "$\\mathbf{w} = (1, 1, 1, 1)$ / standardized",
  "$\\mathbf{w} = (10, 1, 1, 1)$",
  "$\\mathbf{w} = (1, 10, 1, 1)$",
  "$\\mathbf{w} = (1, 1, 10, 1)$",
  "$\\mathbf{w} = (1, 1, 1, 10)$"
)
colnames(final_tab_sim1) <- c(
  "$\\mathit{SSE}$",
  "$\\mathit{SSE_{Var}}$", 
  "$\\mathit{SSE_{Skew}}$",
  "$\\mathit{SSE_{Kurtosis}}$",
  "Average"
)
note <- paste0("The results of Simulation 1 averaged across all ", length(results), " data sets. Cells indicate how well the maximization of the weighted $\\mathit{SSE_{kplusGeneralized}}$ fulfilled each of the four constituting criteria $\\mathit{SSE}$, $\\mathit{SSE_{Var}}$, $\\mathit{SSE_{Skew}}$, and $\\mathit{SSE_{Kurtosis}}$. Cell values represent relative performance, i.e. they were computed as the $\\mathit{SSE}$ that was attained by the given weighting scheme, divided by the best $\\mathit{SSE}$ that was obtained by any weighting scheme in a given simulation run.")

@@

Table 1 illustrates how well each of the six weighting schemes fulfilled each of the four constititung criteria, averaged across the `r length(results)` simulation runs. It is shown that for the weighted approaches, the criterion that received the highest weight was fulfilled to the highest degree, as should be expected. Interestingly, the unweighted approach including standardization had the best average performance across the four criteria. While the advantage over the other weighting schemes does not seem to be remarkable, this pattern was observed in an astonishing `r sum(unstd_best)` of all `r length(results)` simulation runs (`r mean(unstd_best) * 100`%). Among the competing weighting schemes, the unweighted standardization approach therefore seems preferable. It also yielded the best results with regard to maximizing the standard $\mathit{SSE}$. The unweighted standardization approach seemed to implicitly give high priority to minimizing differences in means. This is a pleasant result because, as discussed earlier, for the higher order moment $\mathit{SSE}$s to be valid, the group means have to be close to the overall mean, which is best accomplished if the standard $\mathit{SSE}$ has high priority.

@@

Figure 2 shows that with increasing $N$, k-means and the k-plus methods converged on the same level with regard to minimizing $\Delta \mathit{M}$. This is an important observation because it shows that k-plus anticlustering is capable of addressing multiple objectives at the same time---especially when $N$ increases. However, while k-plus anticlustering was capable of addressing multiple objectives simultaneously, increasing the number of optimization criteria made it more difficult to fulfill each single criterion: generalized k-plus performed worse at minimizing $\Delta \mathit{M}$ and $\Delta \mathit{SD}$ than standard k-plus that only considered the mean and variances. However, as intended, generalized k-plus was best at minimizing $\Delta \mathit{Skew}$ and $\Delta \mathit{Kurtosis}$ and at the same time, this k-plus objective maintained a comparably good level of addressing $\Delta \mathit{M}$ and $\Delta \mathit{SD}$, outperforming diversity anticlustering.

It is maybe surprising that generalized k-plus does not seem to have performed much better than diversity anticlustering with regard to skewness and kurtosis, even though the objective has been specifically tailored to these criteria. Figure 2 gives a more fine grained assessment of this observation: Surprisingly, for small $N$, generalized k-plus was even worse than diversity anticlustering at minimizing $\Delta \mathit{Skew}$ and $\Delta \mathit{Kurtosis}$. However, with increasing $N$, it clearly outperformed diversity anticlustering. These results illustrate the cost of optimizing several criteria at the same time: For low $N$, generalized k-plus was apparently preoccupied with fulfilling the objectives $\Delta \mathit{M}$ and $\Delta \mathit{SD}$. Increasing $N$ then facilitated to also address $\Delta \mathit{Skew}$ and $\Delta \mathit{Kurtosis}$.
